Overlooked Dynamics in Algorithmic Recourse

### 2022-07-06 paper/sections/introduction.qmd
added pseudo-code placeholder

The paper is structured as follows: @sec-related ... .

>>>>

We find that ...

The paper is structured as follows: @sec-related places our work in the broader context of related literature. @sec-method presents our methodology and data. @sec-empirical presents our empirical findings which are then discussed in @sec-discussion. We also point to some of the limitations or our work as well as avenues for future research in @sec-limit. Finally, @sec-conclusion concludes.

### 2022-07-06 paper/sections/methodology.qmd
added pseudo-code placeholder

In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro. Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks. We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
$$
\mathbf{s}\prime = \arg \min_{\mathbf{s}\prime \in \mathcal{S}} \left\{  \ell(M(f(\mathbf{s}\prime)),t)^\mathsf{T} \mathbf{1}_K + \lambda h(f(\mathbf{s}\prime)) \right\}
$$ {#eq-general}

Here $\mathbf{s}\prime$ is a $K$-dimensional vector of counterfactual states and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one. This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared. In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].

Both **REVISE** and **CLUE** search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the deterministic function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards]. CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question $M$. To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The **Greedy** approach proposed by @schut2021generating also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the complexity penalty $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. To this end the authors use a complexity penalty $h(\mathbf{s}\prime)$ that favours diverse outcomes, in the sense that $s_1, ... , s_K$ look as different from each other as possible.

@@

The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of algorithmic recourse in practice. We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class. Note that we focus our attention on classification problems, since classification poses the most common practical use-case for algorithmic recourse. In order to simulate the dynamical process we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$. Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods $T$. To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in ...

[PSEUDO CODE]

@@

$$
MMD^2_u
$$ {#eq-mmd}

@@

The evaluation metric in @eq-mmd is computed after every round $t=1,...,T$ of the experiment. To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution we follow @arcones1992bootstrap. To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$. Then, we split the permuted data into two new samples $X\prime$ and $\tilde{X}\prime$ having the same size as the original samples. Then under the null hypothesis we should have that $MMD_u^2[F,X\prime,\tilde{X}\prime]$ be approximately equal to $MMD_u^2[F,X,\tilde{X}]$. The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

@@

As an alternative to PP MMD we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement). This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured. This is one reason for why PP MMD is our our preferred metric.

>>>>

In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro. Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks. We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:
$$
\mathbf{s}^\prime = \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{  \ell(M(f(\mathbf{s}^\prime)),t)^\mathsf{T} \mathbf{1}_K + \lambda h(f(\mathbf{s}^\prime)) \right\}
$$ {#eq-general}

Here $\mathbf{s}^\prime$ is a $K$-dimensional vector of counterfactual states and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In the case of the baseline counterfactual generator [@wachter2017counterfactual] $f$ is just the idendity function and the number of counterfactuals $K$ is equal to one. This generator, which we shall refer to as **Wachter** in the following, shall serve as the baseline against which all other gradient-based methodologies will be compared. In particular, we include include the following generator in our benchmarking exercises: REVISE [@joshi2019towards], CLUE [@antoran2020getting], DICE [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating].

Both **REVISE** and **CLUE** search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the deterministic function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards]. CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question $M$. To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The **Greedy** approach proposed by @schut2021generating also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the complexity penalty $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Finally, **DICE** distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. To this end the authors use a complexity penalty $h(\mathbf{s}^\prime)$ that favours diverse outcomes, in the sense that $s_1, ... , s_K$ look as different from each other as possible.

@@

The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of algorithmic recourse in practice. We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class. Note that we focus our attention on classification problems, since classification poses the most common practical use-case for algorithmic recourse. In order to simulate the dynamical process we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$. Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods $T$. To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{euclid}

\begin{algorithm}
\caption{PLACEHOLDER}\label{euclid}
\begin{algorithmic}[1]
\Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
\State $r\gets a\bmod b$
\While{$r\not=0$}\Comment{We have the answer if r is 0}
\State $a\gets b$
\State $b\gets r$
\State $r\gets a\bmod b$
\EndWhile\label{euclidendwhile}
\State \textbf{return} $b$\Comment{The gcd is b}
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

$$
\begin{aligned}
MMD^2_u[F,{X}^\prime,\tilde{X}^\prime] &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
\end{aligned}
$$ {#eq-mmd}

@@

The evaluation metric in @eq-mmd is computed after every round $t=1,...,T$ of the experiment. To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution we follow @arcones1992bootstrap. To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$. Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples. Then under the null hypothesis we should have that $MMD_u^2[F,X^\prime,\tilde{X}^\prime]$ be approximately equal to $MMD_u^2[F,X,\tilde{X}]$. The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

@@

As an alternative to PP MMD we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement). This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured. This is one reason for why PP MMD is our our preferred metric.
